Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex
First draft summary of keynotes

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item Frank Loeffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

Arfon Smith of github presented the first keynote. Following points summarize the keynote and the associated discussion:
\begin{enumerate}
    \item Socially driven software development.
    \item Role of software in research.
\end{enumerate}

@@

>>>>

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item Frank Loeffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

\subsection{A recipe for sustainable software, Philip E. Bourne.}

The first keynote
\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
was delievered by Philip E. Bourne of UCSD. The slides
are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}

Philip is a basic biomedical scientist but has formed 4 software
companies. He also helped found PLOS and PDB
\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.

He emphasised that Sustainability for software ``does not just mean more
money from Government.''

He noted that creating effective illustrations is a key function of
software in the biomedical sciences: ``Automating Conceptualisation''
(Visualising molecular structures is a problem that has now been solved,
he is working on 3-d visualisations of the contents of cells.)

He has looked into reproducibility in computational biology, and has
concluded that ``I have proved I cannot reproduce research from my own
lab'':
(Computational
Biology Resources Lack Persistence And Usability, Veretnik, Fink and
Bourne.)\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1000136}}(

He noted that in places, development in science has improved thanks to
open source and Github, but for the most part remains arcane. He argues
that there's a lot we can learn from the app-store model. He
mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and OpenScienceDataCloud
\footnote{\url{https://www.opensciencedatacloud.org/}} as
distribution channels.

He noted a common evolutionary pathway for computational biology
projects, from data archive to analytics platform to educational use. He
suggests that use for outreach might be the final step.

He has problems with staff retention with respect to private organisations which reward 
those combining research and software expertise (the ``Google Bus''), however,
he's a strong fan of private sector sustainability sources and PPP. He
notes that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding requirement
even with a vibrant community -- good will only goes so far if you're
being paid to do something else.

He talked about grant
schemes
 of relevance in the U.S., particularly with regard to technology transfer.  \footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}

He had problems with selling research software: his technology transfer
office wanted wanted huge IP reachthrough, whereby they would get a
share of profits from drugs developed by pharma companies who use the
software. He was aware this was unrealistic but the TTO insisted for a
while. He wants to push a much more one-click approach to purchasing
university-written software.

He then presented the usual arguments on directly valuing software as a
research output alongside papers. Readers will be familiar with this
debate. An interesting reference here is ``What
does peer review mean when applied to computer code?, Rosemary Dickin.''
\footnote{\url{http://blogs.plos.org/biologue/2013/08/08/what-does-peer-review-mean-when-applied-to-computer-code/}}

He then expanded on the idea of the \textbf{digital enterprise}, where
IT underpins the whole of organisational activities. He contends that
universities are way behind the curve on this. In particular, he
highlights the separation of research, teaching and administration into
silos without a common IT framework as a blocker to many useful
organisational innovations: ``University 2.0 is yet to happen.'' He
spoke of a circumstance where someone had used an algorithm developed
for computational biology in marketing.

The role of an institution is important in this space. He argued that
funders can help train institutions not just individuals.

He concluded with a reference to his paper
\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1002001}}``Ten
simple rules for getting ahead as a computational biologist in
academia'' and argued that computational scientists ``have a
responsibility to convince their institutions and communities that
software is scholarship,'' a point with which we strongly agree.

\subsection{Scientifc Software and the Open Collaborative Web}

The second keynote\footnote{\url{http://is.gd/wssspe}} was by
 Arfon Smith\footnote{\url{http://arfon.org/}} of Github.

His talk used an example from data reduction in Astronomy, where one
attempts to remove interfering effects from the device. In this field, a
``bad pixel mask'' is recalculated by each researcher, despite being
persistent over time. He estimated 13 wasted person-years of waste.

He asked, of this practice, ``why didn't we know better?'' and argued
this was because we're taught to focus on immediate research outcomes
and not on continuously improving and building on tools for research. He
then asked, later in our working lives, when we do know better, why
don't we act any different, and argued this was due to incentivisation:
only the immediate products of research, not the software, are valued.

He referenced Victoria Stodden's talk at OKCon
\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},
\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
which he said argued these points well.

He cited the EMCEE \footnote{\url{https://github.com/dfm/emcee}} Markov Chain
Monte Carlo code by Dan Forman Mackey \footnote{\url{http://dan.iel.fm/}} as a
good example of a scientific project using GitHub. Amongst other
examples of good scientific projects on GitHub he used an example from JoeZuntz
\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}.

C. Titus Brown \footnote{\url{http://ged.msu.edu/}} argues that with regard to
reusable software, ``we should just start doing it''. Arfon notes that
documentation should be ``treated as a first class entity.''

He noted that some software fields have excellent cultures of module
reuse, such as rubygems \footnote{\url{http://rubygems.org/}},
PyPI \footnote{\url{https://pypi.python.org/pypi}}, and CPAN
\footnote{\url{http://www.cpan.org/}}, where there is effectively
ultra-low-friction collaboration through these repositories. This has
not happened in highly numerical, compiled language scientific software.

He argues that the GitHub Pull Request code review mechanism facilitates
such collaboration, by allowing one to code first, and seek review and a
merge back into the trunk later.

Fernando Perez says that ``Open Source
is reproducible by necessity'' - since reproducibility is a prerequisite
for remote collaboration.

Arfon therefore defines GitHub as part of the next stage in development
of the web, following on from the ``social web'' of facebook et.al.:
``The Collaborative Web.''

@@

\section{Illustrative use-cases}

Perhaps discuss sustained software use-cases and issues surrounding them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
